Forecasting the 2024 U.S. Presidential Election

### 2024-10-20 paper/paper.qmd
testing git
complete a model with electoral college

```{r}
#| include: false
#| warning: false
#| message: false
# Set a CRAN mirror
# Set a specific CRAN mirror (this example uses the US CRAN mirror)
install.packages("here")

@@

library(tidyverse)
# Install janitor package if not installed
install.packages("janitor")

@@

```{r}
# Build the GLM model
glm_model <- glm(candidate_binary ~ pct + sample_size + methodology + pollscore, 
                 data = data, 
                 family = binomial)

# Summary of the model
summary(glm_model)

@@

```{r}
# Make predictions
predictions <- predict(glm_model, newdata = data, type = "response")

# Convert probabilities into binary outcomes (1 for Harris win, 0 for Trump win)
data$predicted_outcome <- ifelse(predictions > 0.5, 1, 0)

# View the prediction results
head(data)

@@

```{r}
# Confusion matrix
table(data$candidate_binary, data$predicted_outcome)

# Calculate accuracy
accuracy <- mean(data$candidate_binary == data$predicted_outcome)
accuracy

@@

```{r}
# Plot predicted probabilities
ggplot(data, aes(x = pct, y = predictions, color = factor(candidate_binary))) +
  geom_point() +
  labs(title = "Predicted Probability of Harris Winning vs Poll Percentage",
       x = "Polling Percentage", y = "Predicted Probability")

@@

Talk way more about it.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
# Set a CRAN mirror
# Set a specific CRAN mirror (this example uses the US CRAN mirror)

@@

install.packages("here")
library(tidyverse)
# Install janitor package if not installed
install.packages("janitor")

@@

Simple Linear Regression for Kamala Harris and Donald Trump

Kamala Harris
We will perform a simple linear regression using pct (poll percentage) as the dependent variable and sample_size as the independent variable.

```{r}
# Load necessary library
library(ggplot2)

# Filter the data for only "Kamala Harris"
harris_data <- subset(data, candidate_name == "Kamala Harris")

# Perform a simple linear regression with pct as the dependent variable and sample_size as the independent variable
model <- lm(pct ~ sample_size, data = harris_data)

# Summary of the model
summary(model)

# Plot the relationship for Kamala Harris
ggplot(harris_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Linear Regression of Percentage vs Sample Size for Kamala Harris",
       x = "Sample Size",
       y = "Percentage")

@@

Donald Trump
We will repeat the same process for Donald Trump.

```{r}
# Load necessary library
library(ggplot2)

# Filter the data for only "Kamala Harris"
trump_data <- subset(data, candidate_name == "Donald Trump")

# Perform a simple linear regression with pct as the dependent variable and sample_size as the independent variable
model <- lm(pct ~ sample_size, data = trump_data)

# Summary of the model
summary(model)

# Plot the relationship for Kamala Harris
ggplot(trump_data, aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Linear Regression of Percentage vs Sample Size for Donald",
       x = "Sample Size",
       y = "Percentage")

@@

Multiple Linear Regression (MLR) for Kamala Harris and Donald Trump

Kamala Harris
In this step, we will build a multiple linear regression (MLR) model for Kamala Harris using additional predictors like numeric_grade, pollscore, transparency_score, and state.

```{r}
# Convert categorical variables to factors for Kamala Harris
harris_data$state <- as.factor(harris_data$state)
harris_data$methodology <- as.factor(harris_data$methodology)

# Build MLR model for Kamala Harris
mlr_harris_model <- lm(pct ~ numeric_grade + pollscore + transparency_score + sample_size + state + methodology, data = harris_data)

# Summary of the MLR model for Harris
summary(mlr_harris_model)

# Diagnostic plots for the MLR model
par(mfrow = c(2, 2))
plot(mlr_harris_model)
par(mfrow = c(1, 1))  # Reset plot layout

@@

Similarly, we build an MLR model for Donald Trump.
```{r}
# Convert categorical variables to factors for Trump
trump_data$state <- as.factor(trump_data$state)
trump_data$methodology <- as.factor(trump_data$methodology)

# Build MLR model for Donald Trump
mlr_trump_model <- lm(pct ~ numeric_grade + pollscore + transparency_score + sample_size + state + methodology, data = trump_data)

# Summary of the MLR model for Trump
summary(mlr_trump_model)

# Diagnostic plots for the MLR model
par(mfrow = c(2, 2))
plot(mlr_trump_model)
par(mfrow = c(1, 1))  # Reset plot layout
```

Check Multicollinearity using Variance Inflation Factor (VIF)

```{r}
# Load the car package for VIF
if (!require(car)) install.packages("car")
library(car)

# Check VIF for Kamala Harris model
vif(mlr_harris_model)

# Refine the model by removing less significant predictors (e.g., methodology and transparency_score)
mlr_harris_model_refined <- lm(pct ~ numeric_grade + pollscore + sample_size + state, data = harris_data)

# Summary of the refined Harris model
summary(mlr_harris_model_refined)

# Check VIF for the refined model
vif(mlr_harris_model_refined)

# Plot diagnostic plots for the refined model
par(mfrow = c(2, 2))
plot(mlr_harris_model_refined)
par(mfrow = c(1, 1))  # Reset plot layout

@@

Stepwise Model Selection for Optimization

```{r}
# Perform stepwise selection to optimize the Harris model
step_model <- step(mlr_harris_model_refined, direction = "both")

# Summary of the stepwise model
summary(step_model)

# Plot diagnostic plots for the stepwise model
par(mfrow = c(2, 2))
plot(step_model)
par(mfrow = c(1, 1))  # Reset layout

Final Models and Predictions for Kamala Harris and Donald Trump

```{r}
# Final models for Harris and Trump
mlr_harris_model_final <- lm(pct ~ pollscore + log(sample_size) + state, data = harris_data)
mlr_trump_model_final <- lm(pct ~ pollscore + log(sample_size) + state, data = trump_data)

# Predict poll percentages for Kamala Harris
harris_data$predicted_pct_harris <- predict(mlr_harris_model_final, newdata = harris_data)

# Predict poll percentages for Donald Trump
trump_data$predicted_pct_trump <- predict(mlr_trump_model_final, newdata = trump_data)

Electoral College Votes Prediction
In this step, we aggregate the predictions for each state and determine the winner based on the electoral votes.

```{r}
# Define electoral votes for each state (and Washington, D.C.)
electoral_votes <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida", 
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", 
            "Maine CD-1", "Maine CD-2", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
            "Missouri", "Montana", "Nebraska", "Nebraska CD-1", "Nebraska CD-2", "Nebraska CD-3", "Nevada", 
            "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", 
            "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", 
            "District of Columbia"),
  electoral_votes = c(9, 3, 11, 6, 55, 9, 7, 3, 29, 16, 4, 4, 20, 11, 6, 6, 8, 8, 2, 1, 1, 10, 11, 15, 10, 6, 10, 
                      3, 5, 1, 1, 2, 6, 4, 14, 5, 29, 16, 3, 18, 7, 6, 20, 4, 9, 3, 11, 38, 6, 3, 13, 12, 5, 10, 3, 3)
)

# Aggregate predictions by state for Harris and Trump
harris_state_avg <- aggregate(predicted_pct_harris ~ state, data = harris_data, FUN = mean)
trump_state_avg <- aggregate(predicted_pct_trump ~ state, data = trump_data, FUN = mean)

# Merge predictions for both candidates
prediction_comparison <- merge(harris_state_avg, trump_state_avg, by = "state", all.x = TRUE)

# Merge the electoral votes with the predictions
prediction_comparison <- merge(prediction_comparison, electoral_votes, by = "state", all.x = TRUE)

# Determine the winner for each state
prediction_comparison$winner <- ifelse(prediction_comparison$predicted_pct_harris > prediction_comparison$predicted_pct_trump, "Harris", "Trump")

# Calculate total electoral votes for Kamala Harris
harris_electoral_votes <- sum(prediction_comparison$electoral_votes[prediction_comparison$winner == "Harris"], na.rm = TRUE)

# Calculate total electoral votes for Donald Trump
trump_electoral_votes <- sum(prediction_comparison$electoral_votes[prediction_comparison$winner == "Trump"], na.rm = TRUE)

# Print the results
print(paste("Harris Electoral Votes:", harris_electoral_votes))
print(paste("Trump Electoral Votes:", trump_electoral_votes))

# Determine the predicted winner
if (harris_electoral_votes >= 270) {
  print("Kamala Harris is predicted to win the 2024 election.")
} else if (trump_electoral_votes >= 270) {
  print("Donald Trump is predicted to win the 2024 election.")
} else {
  print("No candidate reached 270 electoral votes.")
}

@@

```{r}
# #| echo: false
# #| eval: true
# #| warning: false
# #| message: false

# library(rstanarm)

# first_model <-
#   readRDS(file = here::here("models/first_model.rds"))
```

```{r}
# #| echo: false
# #| eval: true
# #| label: tbl-modelresults
# #| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
# #| warning: false

# modelsummary::modelsummary(
#   list(
#     "First model" = first_model
#   ),
#   statistic = "mad",
#   fmt = 2
# )
```
